---
title: "Getting and Cleaning Data- Course Project Code Book"
author: "Rocio Ana-Sofia Segura"
date: "May 14, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Code Book

##The steps to obtain the analysis and final summary data frames are described below with embedded r code chunks and results to facilitate peer reviews. For a description of the assignment, please see the README.md file. 

### 1. Load libraries
```{r}
library(data.table)
library(dplyr)
```

### 2. Set the working directory
```{r}
setwd("~/coursera/GettingAndCleaningData/data")
getwd()
```

### 3. Import relevant raw data.
(Not in its most raw form but calling it raw data here since this is the most raw form available to me.)
```{r}
xtrain		      <- read.table("/Users/rociosegura/coursera/GettingAndCleaningData/data/UCI HAR Dataset/train/X_train.txt", header = FALSE)
xtest			      <- read.table("/Users/rociosegura/coursera/GettingAndCleaningData/data/UCI HAR Dataset/test/X_test.txt", header = FALSE)
features 	      <- read.table("/Users/rociosegura/coursera/GettingAndCleaningData/data/UCI HAR Dataset/features.txt", header = FALSE)
subject_test 	  <- read.table("/Users/rociosegura/coursera/GettingAndCleaningData/data/UCI HAR Dataset/test/subject_test.txt", header = FALSE)
subject_train 	<- read.table("/Users/rociosegura/coursera/GettingAndCleaningData/data/UCI HAR Dataset/train/subject_train.txt", header = FALSE)
ytrain 	        <- read.table("/Users/rociosegura/coursera/GettingAndCleaningData/data/UCI HAR Dataset/train/y_train.txt", header = FALSE)
ytest 	        <- read.table("/Users/rociosegura/coursera/GettingAndCleaningData/data/UCI HAR Dataset/test/y_test.txt", header = FALSE)
actlabels       <- read.table("/Users/rociosegura/coursera/GettingAndCleaningData/data/UCI HAR Dataset/activity_labels.txt", header = FALSE)
```

Join activity labels to ytrain and ytest data frames
```{r}
ytrain <- merge(ytrain, actlabels, by = "V1")
ytrain <- setnames(ytrain, old='V2', new = 'activity')

ytest <- merge(ytest, actlabels, by = "V1")
ytest <- setnames(ytest, old='V2', new = 'activity')
```

### 4. Examine the data
Here I created a function to run str on all the data I imported in a for loop.
```{r}
dflist <- list(xtrain, xtest, features, subject_test, subject_train, ytrain,ytest, actlabels)
dfnames <- list("xtrain","xtest","features","subject_test", "subject_train","ytrain","ytest","actlabels")

mystrfxn <- function(x,y) {
  
  for(i in 1:length(x)) {
    print(paste('Data Frame Name: ' , y[i]))
    str(x[i])
  }
  return
}

mystrfxn(x = dflist, y = dfnames)
```

### 5. Generate valid column names 
Here I chose to use the variable names already provided with the UCI HAR Datasets. 
* Get column names from features data frame and save to new vector named features2
* Process names so that variables are human readable
```{r}
features2 <- mutate(features, tmpvarname = gsub("\\(\\)","", V2), newvarnames = gsub("-","_", tmpvarname))
```
### Select the variable (newvarnames) that corresponds to the processed variable names
```{r}
varnames <- as.vector(t(select(features2, newvarnames)))
```

* There are values (vector of column names) in varnames that are duplicated
* Use make.names with unique equal to TRUE so that all column names get a unique value.
* Save to a new vector named validvarnames
```{r}
validvarnames <-  make.names(varnames, unique = TRUE)
```

* Use validvarnames vector as the source for the new names
```{r}
oldtrainnames <- names(xtrain)
xtrain <- setnames(xtrain, old = oldtrainnames, new = validvarnames)

oldtestnames <- names(xtest)
xtest <- setnames(xtest, old = oldtestnames, new = validvarnames)
```


### 6. Add new columns, source, subjectid, and activity

* Add a new column to identify whether test or train data
```{r}
xtrain  <- mutate(xtrain, source="train")
xtest   <- mutate(xtest, source="test")
```

* Add a subject id and activity column from subject_test/train and ytest/xtest respectively
```{r}
xtrain <- cbind(subjectid=subject_train$V1, activity = ytrain$activity, xtrain)
xtest <- cbind(subjectid=subject_test$V1, activity = ytest$activity, xtest)
```



### 7. Merge x_train.txt and x_test.txt using rbind()
```{r}
train_test_appended <- rbind(xtrain, xtest) 
```



* Arrange subjectid and source so they appear as the first columns
```{r}
train_test_appended <- train_test_appended %>%
  select(source, everything())
train_test_appended <- train_test_appended %>%
  select(subjectid, everything())
```



### 8. Use select to find all column names that contain "mean_", "std_", "source", "subjectid", or "activity"
```{r}
acc_mean_std_analset <- train_test_appended %>%
  select(matches('mean_|std_|source|subjectid|activity')) 
```

### Analysis data frame - (acc_mean_std_analset)
The measurements included in this data frame are limited to the mean and standard deviation variables plus 3 extra columns that are helpful in summarizing the data. 

* subjectid (subject id number from the subject_test/train txt input files)
* source (new column I created to denote whether subject was either in the train or test group)
* activity (column added from the ytest/ytrain txt input files and activity labels joined from the activity_labels.txt file)
```{r}
str(acc_mean_std_analset)
head(acc_mean_std_analset, 5)
```

### 9. Summarize
The final data frame, acc_mean_std_avg, contains the averages for the means and standard deviation measures grouped by subjectid, source and activity.
```{r}
acc_mean_std_avg <- acc_mean_std_analset %>% group_by(subjectid, source, activity) %>% summarise_each(funs(mean))
str(acc_mean_std_avg)
head(acc_mean_std_avg, 5)
```

### 10. Output summary dataset to txt file
use write.table() using row.name=FALSE
```{r}
write.table(acc_mean_std_avg, file = 'acc_mean_std_avg.txt', row.names = FALSE, sep = "\t")
```





